VFTS682: a confirmed dynamical ejection?

### 2018-05-08 VFTS682.tex
started writing the paper
polishing, hardcore stuff still to be done

\institute{{Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands}
  \and{School of Astronomy \& Space Science, University of the Chinese
    Academy of Sciences, Beijing 100012, China}
  \and{National Astronomical Observatories, Chinese Academy of
    Sciences, Beijing 100012, China}
}

@@

\keywords{stars: kinematics, stars:individual:VFTS682}
\maketitle{}

@@

How do stars form is one longstanding question in astrophysics \cite{lada:03}
\todo{ref?}. Answering it for massive stars is particularly difficult,
since these are intrinsically rare
\citep[e.g.,][]{salpeter:55,kroupa:01, schneider:18}, evolve fast, and
remain enshrouded in their parent cloud during the whole formation
process \todo{ref}. Moreover, observations of young massive stars reveal a
complicated multipliticy structure which requires
explanation \citep[][]{kobulnicky:07,mason:09,sana:11,sana:12,kiminki:12,chini:12,kobulnicky:14,almeida:17,demarco:17}. Understanding massive star formation, possibly as a
function of metallicity, is a key question given the present and upcoming
transient survey \citep[e.g., LSST, BlackGem, LIGO/Virgo O3][]{} which
will reveal transients associated to massive stars
evolution and death.

\todo{short summary of formation scenarios goes here}.

The second data release (DR2) from the Gaia satellite
\cite[][]{brown:18,} allows us to test
these hypothesis using one particular star, VFTS682. This star is a
very massive ($\sim$$150\,M_\odot$) WNh5 star in the 30 Doradus region
of the Large Magellanic Cloud, and it is presently observed at a
projected distance of $\sim$$29$\,pc from the nearest cluster of
massive stars R136 \citep[][]{bestenlehner:11}. Based on the extremely high mass of this star and
its present day apparent isolation, \cite{bestenlehner:11} proposed it
might be a candidate for isolated star formation, or a runaway ejected
from R136 in the past.

Massive stars can in principle be ejected from R136 as a consequence of dynamical
interactions \citep[][]{poveda:67,leonard:91, evans:10, fujii:11,
  allison:12, oh:16}, or by the disruption of a binary by the first
core-collapse supernova
\citep[][]{zwicky:57,blaauw:61,dedonder:97,eldridge:11,renzo:18}. However,
R136 has an estimated age of $\lessim2$\,Myr \todo{ref}, which is
shorter than the shortest stellar lifetime of about $\sim$3\,Myr
\citep[e.g.,][]{zapartas:17}, so one would not expect the binary
disruption scenario to be relevant for this cluster.

@@

In \Secref{sec:sample}, we describe how we select stars in R136 and in
the surroundings of VFTS682 itself to remove the mean motion of the
region. \Secref{sec:results} presents our main findings. We conclude by
discussing the implications for theories of
star formation, N-body interactions, and binary evolution in
\Secref{sec:discussion}.

\section{Gaia DR2 data selection}
\label{sec:sample}
VFTS682 is labeled in the Gaia DR2
catalog\footnote{\url{https://gea.esac.esa.int/archive/}} with the
source id 4657685637907503744. For this star, we retrieve from the
Gaia catalog its position in right ascension (RA) and declination (DE)
in the IGCS frame\cite[][]{brown:18}\todo{some 2016 Gaia paper?}, its
proper motion components ($\mu_\mathrm{RA}$, and $\mu_\mathrm{DE}$,
respectively). For the radial velocity of VFTS682 and of the 30 Doradus
region as a whole, we instead use the VFTS data
as quoted in \cite{bestenlehner:11}. Our main conclusions do not
depend on the detailed value of the radial velocity.\todo{improve
  phrasing} \Tabref{tab:vfts682} lists the values adopted throughout
this work for each of these quantities.

@@

To compare the astrometry of VFTS682 and derive its peculiar motion,
we then select data from the Gaia DR2 catalog for two regions: the
``surroundings'' of VFTS 682, and the ``R136 cluster''. The
surrounding region is defined by all the stars within 10 arcminutes from
VFTS682. This includes the subset of stars which we consider to be
part of R136. We include all the stars for which a value of the proper
motion components and the corresponding errors are available.
The ``R136 cluster'' is effectively defined by taking all the stars
within 25 arcseconds from R136a, one of the most massive members of
the cluster itself \citep[][]{crowther:10}.

@@

The 
\section{The kinematics of VFTS682}
\label{sec:results}

@@

\section{Discussion}
\label{sec:discussion}

>>>>

\institute{{Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands}
  % \and{School of Astronomy \& Space Science, University of the Chinese
  %   Academy of Sciences, Beijing 100012, China}
  % \and{National Astronomical Observatories, Chinese Academy of
  %   Sciences, Beijing 100012, China}
}

@@

\keywords{stars: kinematics, stars: runaways, stars: individual: VFTS682}
\maketitle{}

@@

How do stars form is one longstanding question in astrophysics
\cite{lada:03, zinnecker:07}. It is particularly difficult for massive stars, because these are intrinsically rare
\citep[e.g.,][]{salpeter:55,kroupa:01, schneider:18}, evolve fast, and
remain enshrouded in their parent cloud during the formation
process. Moreover, observations of young massive stars reveal a
complicated multipliticy structure which requires
explanation \citep[][]{kobulnicky:07,mason:09,sana:11,sana:12,kiminki:12,chini:12,kobulnicky:14,almeida:17,demarco:17}. Understanding massive star formation, possibly as a
function of metallicity, is a key question given the present and upcoming
transient survey \citep[e.g., LSST, BlackGem, LIGO/Virgo O3][]{} which
will reveal transients associated to massive stars
evolution and death.

The second data release (DR2) from the Gaia satellite
\cite[][]{gaia:16,brown:18} allows us to test
these hypothesis using one particular star, VFTS682. 
This star is a
very massive \citep[$M_\mathrm{ZAMS}\simeq150\,M_\odot$,][]{bestenlehner:11,schneider:18} WNh5 star in the 30 Doradus region
of the Large Magellanic Cloud (LMC), and it is presently observed at a
projected distance of $\sim$$29$\,pc from the nearest cluster of
massive stars R136 \citep[][]{bestenlehner:11}. Based on the extremely high mass of this star and
its present day apparent isolation, \cite{bestenlehner:11} proposed it
might be a candidate for isolated star formation, or a ``slow runaway'' ejected
from R136 in the past. This second option is also supported by the
N-body simulations of \cite{fujii:11, banerjee:12}.   Many other
very massive stars are present in the surroundings of R136, and a more
detailed analysis on the larger sample is desirable.

Massive stars can in principle be ejected from R136 as a consequence of dynamical
interactions \citep[][]{poveda:67,leonard:91, evans:10, fujii:11,
  allison:12, oh:16}, or by the disruption of a binary by the first
core-collapse supernova
\citep[][]{zwicky:57,blaauw:61,dedonder:97,eldridge:11,renzo:18}. However,
R136 has an estimated age of $\lesssim2$\,Myr \citep[][]{sabbi:12}, which is
shorter than the shortest stellar lifetime
\citep[$\sim$3\,Myr, e.g.,][]{zapartas:17}, so one would not expect the binary
disruption scenario to be relevant for this cluster.

@@

In \Secref{sec:sample}, we describe the data selection and validation
process for VFTS 682 and for stars in the surrounding used to
calculate and remove the mean motion of the
region. \Secref{sec:results} presents our main findings. We conclude
by discussing the implications for theories of
star formation, N-body interactions, and binary evolution in
\Secref{sec:discussion}.

\section{Gaia DR2 data selection}
\label{sec:sample}

VFTS682 is labeled in the Gaia DR2
catalog\footnote{\url{https://gea.esac.esa.int/archive/}} with the
source id 4657685637907503744. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution \citep[][]{lindengren:18}. Its G-band
magnitude is 15.65, cf. the V-band magnitude of 16.08
\citep[][]{evans:11, bestenlehner:11}, and the reported
\texttt{astrometric\_noise} = 0. These values suggest that the Gaia
data for VFTS682 are trustworthy. However, the effective temperature
reported is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax value reported in the Gaia catalog.

We retrieve for VFTS682 the position in right ascension (RA) and declination (DE)
in the IGCS frame \cite[][]{brown:18}, its
proper motion components ($\mu_\mathrm{RA}$, and $\mu_\mathrm{DE}$,
respectively). For the radial velocity of VFTS682 and of the 30 Doradus
region as a whole, we instead use the VFTS data
as quoted in \cite{bestenlehner:11}. \Tabref{tab:vfts682} lists the values adopted throughout
this work for each of these quantities.

@@

To compare the astrometry of VFTS682 and derive its peculiar motion,
we then select data from the Gaia DR2 catalog for two regions: the
``surroundings'' of VFTS 682, and the ``R136 cluster''. The
surrounding region is defined by all the stars within 10 arcminutes from
VFTS682 \todo{without a pm and with G mag<17 for VFTS completeness and
to focus on massive stars only}. This includes the subset of stars which we consider to be
part of R136. We include all the stars for which a value of the proper
motion components and the corresponding errors are available.
The ``R136 cluster'' is effectively defined by taking all the stars
within 25 arcseconds from R136a, one of the most massive members of
the cluster itself \citep[][]{crowther:10}. \todo{improve definition
  of stars from R136 -- check with Danny}

@@

The data retrieved, and the ipython notebook used for the analysis
presented here will be made available at \todo{probably git repo on bitbucket?}. 
\section{The kinematics of VFTS682}
\label{sec:results}

@@

\section{Summary and Discussion}
\label{sec:discussion}

\begin{itemize}
\item VFTS682 is a bona fide runaway with $v\sim60\,\mathrm{km\
    s^{-1}}$ thrown out from R136. Both its speed and the age of the
  cluster are consitent with a dynamical ejection.
  \item VFTS682 comes from R136 as was expected by
  \cite{bestenlehner:11, fujii:11, banerjee:12}, so it does not
  require isolated SFH to be explained
\item apparent age tension (connect to VFTS16 as well).
\item is R136 a single young cluster or a merger
\item estimate the influence of the gravitational potential of R136,
  what is its total mass and relaxation time?
\end{itemize}

Random notes: vsini<200 form \cite{schneider:18}, age $1\pm 0.2$\,Myr
from \cite{schneider:18}
